Hybrid platforms and bargaining power

### 2024-01-12 src/paper/application.tex
Expand intro

This paper aims to contribute to this discussion by examining an important, but so far underexplored aspect of platforms: the bargaining power disparity between them and the other market participants, and how this disparity is affected by the platforms' hybrid operation.
In particular, I propose an analytically tractable framework which captures many important aspects of bargaining between one large and a continuum of small players, and apply it to the setting of a hybrid platform facing a continuum of potential entrants.
The results show that, in the presence of bargaining, hybrid operation can indeed have negative welfare consequences for consumers.
This result is particularly striking in this otherwise distortion-free\footnote{
    In the benchmark model without bargaining, the platform's hybrid operation does not affect consumer welfare.
} setting. 
The intuition is that hybrid operation increases the platform's bargaining power against the sntrant sellers.
This in turn leads to a higher entry fee, fewer entrants, then results in lower consumer surplus.
These observations constitute a so-far overlooked aspect of hybrid platforms that policymakers should be aware of.

The closest paper to this one is \textcite{anderson2021hybrid}, which utilizes a similar model and reaches similar conclusions, although through a different mechanism.
I adapt many of the main features of their model with a couple of crucial differences.
Instead of the platform setting a percentage entry fee and being able to commit to it, I assume that (1) platforms charge a lump-sum entry fee and that (2) this entry fee is the result of a negotiation process\footnote{
    The bargaining process is modeled using a solution concept from cooperative game theory, namely the Shapley-value.
    This allows for a tractable analysis while at the same time capturing many of the important features of bargaining.
    Examples of this approach in the industrial organization literature include \textcite{montez2007downstream}, as well as \textcite{hart1990property}, \textcite{levy1997individual}, \textcite{inderst2003bargaining} and \textcite{brugemann2019intra} among others.
}
between the platform and the entrants.
The second point is a novelty not only compared to \textcite{anderson2021hybrid}, but also to the existing literature on hybrid platforms.
Furthermore, in contrast to \textcite{hagiu2022should}, this model is free from any distortive behavior, such as self-preferencing or steering, and the negative results are purely driven by changes in the platform's bargaining position.

This paper is also related to various other strands of the industrial organization literature.
For one, it belongs to the research agenda on understanding platforms and their role in the economy \parencite[e.g.][]{rochet2003platform,hagiu2004optimal,armstrong2006competition,evans2011platform,lee2014competing}.
Furthermore, it is somewhat adjacent to the literature on the importance of exclusive content \parencite[e.g.][]{hagiu2011exclusivity,lee2013vertical,dou2014sell,weeds2016tv}, with the difference that instead of exclusive content giving an advantage against competing platforms, in this paper, own products provide an advantage over entrants.
More generally, many results and concepts of this paper are also applicable outside the context of (multi-sided) platforms.
Certain structures, such as vertical relationships, or even traditional retail stores share a number of features with platforms (notably, the presence of a dominant, indispensable entity), and can be modeled in much the same way.
Therefore, the methodology and results in this paper may also be of interest to researchers studying questions such as retailers having their own private labels \parencite{steiner2004nature} or vertical integration in upstream-downstream markets \parencite{hart1990vertical,aghion2006vertical}.
\textcite{de2005vertical} and \textcite{montez2007downstream}, in which bargaining between the upstream and downstream firms take the center stage, are particularly closely related.

>>>>

This paper aims to contribute to this discussion by examining an important, but so far underexplored aspect of platforms: the bargaining power disparity between them and the other market participants, and how this disparity is affected by the platforms' hybrid operation.
In particular, I propose an analytically tractable framework which captures many important aspects of bargaining between one large and a continuum of small players, and apply it to the setting of a hybrid platform facing a continuum of potential entrants.
The framework quite general, and does not depend on a specific demand structure, investment cost assumptions, and there is even some flexibility to the assumptions about bargaining outcomes.
Nevertheless, even in this stylized setting, I can highlight certain factors which might cause a platform's hybrid operation to potentially have negative effects. 
In particular, when the platform's and the potential entrants' products are substitutes, increasing the platform's product variety will lead to a decrease in the number of entrants.
Under slightly more restrictive assumptions about the functional form of the profit functions, I show that this decrease in entry is so large as to cause a decrease in total profits, as well.
Due to the generality of this model, these results are applicable not only to hybrid platform, but to other similar settings, such as upstream producers having their own downstream outlets, or retailers with private labels, as well.

As a more specific example, I also examine this phenomenon in a more fully specified model, too.
In particular, I provide some sharper results under the assumption of logit-like demand and a specific form of the bargaiing rule.
I show that, in the presence of bargaining, hybrid operation can indeed have negative welfare consequences for consumers.
This result is particularly striking in this otherwise distortion-free\footnote{
    In the benchmark model without bargaining, the platform's hybrid operation does not affect consumer welfare.
} setting. 
The intuition is that hybrid operation increases the platform's bargaining power against the entrant sellers.
This in turn leads to a higher entry fee, fewer entrants, then results in lower consumer surplus.
These observations constitute a so-far overlooked aspect of hybrid platforms that policymakers should be aware of.

The closest paper to this second, more applied part of this one is \textcite{anderson2021hybrid}.
The main difference between the two boils down to how the entry fee is decided.
In the former, of the platform setting a percentage entry fee and being able to commit to it, I assume that (1) platforms charge a lump-sum entry fee and that (2) this entry fee is the result of a negotiation process\footnote{
    The bargaining process is modeled using a solution concept from cooperative game theory, namely the Shapley-value.
    This allows for a tractable analysis while at the same time capturing many of the important features of bargaining.
    Examples of this approach in the industrial organization literature include \textcite{montez2007downstream}, as well as \textcite{hart1990property}, \textcite{levy1997individual}, \textcite{inderst2003bargaining} and \textcite{brugemann2019intra} among others.
}
between the platform and the entrants.
The second point is a novelty not only compared to \textcite{anderson2021hybrid}, but also to the existing literature on hybrid platforms.
Furthermore, in contrast to \textcite{hagiu2022should}, this model is free from any distortive behavior, such as self-preferencing or steering, and the negative results are purely driven by changes in the platform's bargaining position.

The first set of results in this paper go beyond the aforementined studies in the sense that they illuminate the effects of hybridness as a more abstract phenomenon, and thus also highlight the similarities between the hybrid platform literature and adjacent ones.
Certain structures, such as vertical relationships, or even traditional retail stores share a number of features with platforms (notably, the presence of a dominant, indispensable entity), and can be modeled in much the same way.
Therefore, the methodology and results in this paper may also be of interest to researchers studying questions such as retailers having their own private labels \parencite{steiner2004nature} or vertical integration in upstream-downstream markets \parencite{hart1990vertical,aghion2006vertical}.
\textcite{de2005vertical} and \textcite{montez2007downstream}, in which bargaining between the upstream and downstream firms take the center stage, are particularly closely related.
Among empirical studies in related settings, \textcite{ho2017insurer} looks at the effect of insurer competition on health care prices, while \textcite{crawford2018welfare} explores the effects of vertical integration on television markets.

This paper is also related to various other strands of the industrial organization literature.
For one, it belongs to the research agenda on understanding platforms and their role in the economy \parencite[e.g.][]{rochet2003platform,hagiu2004optimal,armstrong2006competition,evans2011platform,lee2014competing}.
Furthermore, it is somewhat adjacent to the literature on the importance of exclusive content \parencite[e.g.][]{hagiu2011exclusivity,lee2013vertical,dou2014sell,weeds2016tv}, with the difference that instead of exclusive content giving an advantage against competing platforms, in this paper, own products provide an advantage over potential entrants.
More generally, many results and concepts of this paper are also applicable outside the context of (multi-sided) platforms.
